---
title: "Chi-squared tests for unfair decks"
keywords: "Probability"
date: 2026-09-27
format: html
---

## Modeling cards and shuffled decks

### Decks as vectors

We followed online code to produce
```{r do1}
suits <- c("Spades", "Hearts", "Diamonds", "Clubs")
ranks <- c("Ace", 2:10, "Jack", "Queen", "King")
# Combine ranks and suits to make 52 unique cards
deck <- paste(rep(ranks, 4), rep(suits, each = 13))
```
A deck is a vector with 52 elements.
This vector "models" a deck of playing cards.

## Draws as samples

With R's `sample` function, we can model shuffling.
We use `set.seed` to obtain a reproducible random
number generation state.

```{r lkshuf}
top_card = function(x) { 
   stopifnot(length(x)==52, is.atomic(x))
   sample(x, size=1)
}
set.seed(1234)
drs = replicate(2000, top_card(deck))
head(table(drs))
mean(table(drs))
2000/52
```

## A model for fairness

### Intuitions

For a fair deck with thorough shuffling and replacement, we 
have an intuition that each
card type has the same chance of being drawn as the top card.

In 2000 draws, the _expected_ number of appearances of "2 Hearts"
is 2000/52, and that is the same for all card types.

### Formalism

Formally, we can abstract from the actual names of the cards and
represent the shuffling/drawing procedure in this way:

Let $i = 1, \ldots, 52$ denote card types

Let $j = 1, \ldots, 2000$ denote shuffle/draw/replace events

Define

$$
C_{ij} =
\begin{cases}
  1 & \text{if draw } j \text{ yields card type } i, \\
  0 & \text{otherwise.}
\end{cases}
$$

Now define $X_i = \sum_j C_{ij}$.  Our statement above about the
intuitive expectation on the number of appearances of any given
card type will be written
$$
EX_i = 2000/52 \text{ for all } i = 1,\ldots, 52.
$$

This is equivalent to the following formulation.  Let $p_i, i=1, \ldots, 52$
denote the _probability_ (not yet defined in any formal way) that our shuffle/draw procedure
yields card of type $i$.  Then $EX_i = 2000 p_i$ for all $i$.  For the
general case of $n$ draws, $EX_i = n p_i$.

## Measuring departure from fairness

A well-established metric for assessing departure from fairness is
called "chi-squared".  We will use the symbol $D$ to represent this
measure.  Large values of $D$ can be indicative of departure of the
observed frequencies from the expected values.

$$
D = \sum_{i=1}^{52} \frac{(X_i - EX_i)^2}{EX_i}
$$

We'll implement $D$ as `mystat` and show that it is equivalent to
a quantity computed in R with `chisq.test`.

```{r OmE}
mystat = function(dat, expected) sum((table(dat)-expected)^2/expected)
ExpectedCount = 2000/52
sum((table(drs)-ExpectedCount)^2/ExpectedCount)
chisq.test(table(drs))
mystat(drs, ExpectedCount)
```

Note that `chisq.test` reports the same statistic as `mystat`, but
includes values for `df` (degrees of freedom) and `p-value`.  "Degrees of freedom"
refers here to the fact that we have a fixed number of count categories (52 card types),
and if 51 are known, the value for the 52nd can be deduced -- it is "unfree".

The p-value is a quantity that, when small, can be indicative of
a departure from the expected outcome of an experiment.  We will
return to that concept at a later date.

## Replicating our test of fairness many times

Any given collection of shuffle/draw events will yield
a value of $D$.  Different series will yield different
values of $D$, but when the deck is fair, the variation
in $D$ is prescribed by the mathematics of probability.

We will make a histogram of 100 replicates of 2000 draws
to exhibit variation in $D$.

```{r domore}
set.seed(1234)
predist = lapply(1:100, function(x) replicate(2000, top_card(deck)))
dist = sapply(predist, function(x)mystat(x, expected=2000/52))
hist(dist, prob=TRUE)
```

With 1000 replicates, we get a more pleasing display.

```{r domore2, cache=TRUE}
set.seed(1234)
predist = lapply(1:1000, function(x) replicate(2000, top_card(deck)))
dist = sapply(predist, function(x)mystat(x, expected=2000/52))
hist(dist, probability=TRUE)
```

The form of this histogram (drawn in R with the `probability` argument set to TRUE)
can be approximated by the density of the chi-squared distribution
with 51 degrees of freedom.

```{r lkdens}
hist(dist, probability=TRUE)
lines(1:100, dchisq(1:100, 51))
```

The "51" fixes the degrees of freedom of the model for which
the density will be drawn.

**Nota bene.**  The fact that we can approximate the shape of the
histogram for this fairly complicated data generation process,
using the 'density function' for the chi-squared probability
model with specified degrees of freedom, is a minor miracle
building on almost two centuries of mathematical, statistical,
and computational research and development.

We might as well have a look at the density function for the
chi-squared distribution with $k$ degrees of freedom:

$$
f(x; k) =
\begin{cases}
  \dfrac{1}{2^{k/2}\,\Gamma(k/2)}\, x^{k/2 - 1} e^{-x/2}, & x > 0, \\[1ex]
  0, & \text{otherwise},
\end{cases}
\qquad k = 1, 2, \ldots
$$

In our call to `dchisq` we evaluated this function at points $x = 1, 2, ..., 100$
with $k = 51$.  $e$ is Euler's constant `r round(exp(1),5)`... 
$= \lim_{n \to \infty} \left(1 + \frac{1}{n}\right)^{n}$ , and $\Gamma$
denotes the "Gamma function" defined as $\Gamma(x) = \int\limits_{0}^{\infty} e^{-t}t^{x-1}dt$
which satisfies $\Gamma(x+1) = x\Gamma(x)$.

## Interpret


What's the difference between these two runs?

```{r lkbad}
set.seed(1234)
baddeck = deck
baddeck[1] = baddeck[52]
newdrs = replicate(2000, top_card(baddeck))
mystat(newdrs, ExpectedCount)
```

```{r lkbad2}
set.seed(1234)
baddeck[2] = baddeck[51]
newdrs = replicate(2000, top_card(baddeck))
mystat(newdrs, ExpectedCount)
```

How is this different?


```{r lkbad3}
set.seed(1234)
baddeck[2] = baddeck[52]
newdrs = replicate(2000, top_card(baddeck))
mystat(newdrs, ExpectedCount)
```

What's wrong here?

```{r lkbad4}
set.seed(1234)
baddeck[2] = baddeck[52]
newdrs = replicate(1000, top_card(baddeck))
mystat(newdrs, ExpectedCount)
```

## CAVEAT

Our theory of the distribution of $D$ depends on a few
assumptions.  The most important one is that the events
underlying contributions to $D$ are _independent_.

Establishing independence in real-world contexts is
often very difficult.  One must be very vigilant,
searching for hidden sources of dependence between
events emerging from measurement processes.

## EXERCISE

Go through this document changing the random process from
sampling from a deck of playing cards to rolling a fair
or biased 6-sided die.
